---
title: "Relation between OER adoption and persistence in higher education"
output: pdf_document
fontsize: 12
bibliography: r-references.bib
csl: apa.csl
link-citations: true
---

```{r globalOpts, echo=FALSE, warning=FALSE, message=FALSE, comment=""}
knitr::opts_chunk$set(fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='center', cache=FALSE, progress=TRUE)
```


```{r setup, echo=FALSE, warning=FALSE, message=FALSE, comment=""}
# Source supporting files
source("~/Documents/oer-history/supportingScripts/descriptives.R")
```


```{r f17 figTbl}
figs <- captioner(prefix="Figure ", auto_space = FALSE) 
tbls <- captioner(prefix="Table ", auto_space = FALSE)
# to reference fig in text, use the following syntax: 
# `r figs("Logic4", display = "cite")` #Where "Logic4" is the name of the chunk that produces the image
# `r tbls("bioModels4",display = "cite")`

#options, 3 digits, hide all warnings
options(digits = 3, warn = -1, knitr.kable.NA = " ")
```



<!--# Introduction

## Setting up the problem:Rising textbook cost

  A major problem currently facing education in America is the rising cost of obtaining a college degree. While tuition hikes predominate student costs, an additional source of expense is from textbook prices that have been found to rise at an alarming rate. While tuition and fees have increased about 63% from January 2006 to December 2017 [@bls2016], textbook prices have increased by about 88% (not seasonally adjusted; [@blsbk]). This meteoric rise in textbook costs drastically influences students' academic choices [@FloridaVirtualCampus2016; @Senack2014]. A survey of over 22,000 students in Florida's higher education institutions revealed that exorbitant textbook costs often determine whether students will forego purchasing a textbook (67%), enroll in fewer class (48%), drop (26%) or withdraw (21%) from a class, earn a poor grade because they were unable to buy a book (38%) [@FloridaVirtualCampus2016]. Similar patterns of responses were found in another survey of over 2000 students across 150 American institutions [@Senack2014]. Senack [-@Senack2014] found that 65% of the students surveyed avoided purchasing an expensive textbook, despite being concerned that it would negatively impact their course performance. Senack [-@Senack2014] also found that 82% of the students believed that they would perform better in their courses if their textbook was freely available online and purchasing the printed copy was optional.

## OER as a solution to the problem, and open textbook as a type of OER

  In order to mitigate the burdens of rising textbook costs on students, there has been a surge of interest in open educational resources (OER). OER differs from traditional educational materials in that they are freely available online. OERs include open-textbooks (e.g. OpenStax), lecture videos (e.g. Khan Academy), simulations (e.g. PhET interactive simulations), complete courses (e.g. MIT OpenCourseWare) to name a few. A recent report released by the Babson survey found that while awareness about OER amongst faculty is low (~30%), yet there is a steady increase in OER adoption [@Seaman2017]. In this paper, we will focus on open textbooks as a form of OER. In addition to being freely available, open textbooks can be immediately accessed, and are often governed by a Creative Commons license that legally allows users to alter the materials and redistribute it, thereby ensuring flexibility in the academic process. A recent survey of 2,711 faculty across US insitutions found that, adoption of open textbooks increased from 5% in 2015-2016 to 9% in 2016-2017, saving students anywhere between \$68 to \$182 [@Seaman2017]. 
  
  While open-textbooks have undoubtedly been successful at reducing student costs, the critical question is whether these cost savings translate into improved student academic outcomes. One fundamental way in which OER has the ability to help students' academic outcomes is by ensuring that students have access to the course materials at the very outset of a course. However, having access to course materials can only result in changes in academic outcomes when students, who would not have otherwise been able to afford to buy an expensive textbook, engage in using the OER that is available to them. If these students do not take advantage of this resource, we can assume that their academic outcomes will remain similar to if they did not have the course textbook.
  
  In this project we focus on open textbook as a type of OER. While open textbooks are freely available to students online, they often have printed versions available at a very small cost. OER is presumed to help in this regard by improving access to beneficial learning materials, however research in this area is has found equivocal results (Hilton, 2016). 


## Difference between open and traditional textbook

## Going beyond access, OER's impact on academic outcomes
### Prior research
## Factors affecting academic outcomes other than OER
### Demographic factors
### Prior achievement 
### course characteristics


## Goals of the project-->

# Methods

In order to evaluate the impact of open textbook adoption on academic outcomes of an introductory history course at SLCC, we obtained the following student records from the period of Fall 2011 to Fall 2017 from the co-authors at SLCC:   

 **_course metadata_** (e.g., instructor id, course number, section id, term start and end dates)       
 **_student characteristics_** (e.g., ethnicity, gender, pell eligibility, registration status)        
 **_prior achievement scores_** (e.g., cumulative undergraduate GPA, college placement test scores) and,            
 **_course performance_** (e.g., final course grade, credits earned)           

Furthermore, this dataset included the above parameters for 2 other social sciences subjects (political science, economics) that exclusively used conventional textbooks during the same period of time. All the posthoc analyses of the data that ensued, used the data from the 2 comparable social sciences subjects in addition to history in order to tease apart any temporal fluctuations in student performance that would have occurred even without the use of open textbooks. 

Besides these student records, we also obtained the history course syllabi for some sections starting from Fall 2015, to Fall 2017. We used the syllabi data in order to extract textbook names and general course structure for the history courses using traditional and open-textbooks.   

## Participants

```{r participantTable}
participantTable %>% 
  knitr::kable(format = "latex", linesep = "", valign = "t", booktabs = TRUE, caption = "The number of student records (n) in each subject and textbook type, alongwith their pell eligibility status.", col.names = c("Pell Eligibility", "n"), align = c("l", "r")) %>% 
  #collapse_rows() %>% 
  kable_styling(full_width = FALSE, font_size = 12) %>% 
  group_rows("History Open Textbook", start_row = 1, end_row = 2) %>% 
  group_rows("History Traditional Textbook", start_row = 3, end_row = 4) %>% 
  group_rows("Economics Traditional Textbook", start_row = 5, end_row = 6) %>% 
  group_rows("Political Science Traditional Textbook", start_row = 7, end_row = 8)  
```

We examined the student records included in the dataset that we obtained from SLCC. After removing records of students who were less than 18 years of age at the time of the course, we had a total of `r dim(SLCC_11_18_proc)[1]` student records in the dataset. We observed that there were `r n_distinct(SLCC_11_18_proc$studentId)` unique students in the data. Indeed, out of all the records, `r sum(duplicated(SLCC_11_18_proc$studentId))` records belonged to students who had either taken more than one of the social science courses in the dataset, or had taken a course more than once, or both. The sample was comprised of students with a mean age of `r round(mean(SLCC_11_18_proc$age, na.rm = TRUE))`, out of whom `r gender$genderPerc[1]`% were female, `r gender$genderPerc[2]`% were male students, and the remaining students chose not to report their gender. Out of these students, `r pell$pell[pell$everPellEligibleInd == "Y"]`% were pell eligible, `r regStatus$enroll[regStatus$fullTime == "Full-Time"]`% were in full-time programs, and `r firstGen$firstGen[firstGen$firstGenerationInd == "Y"]`% identified as first generation students. 

## Procedure

Studies of OER that have evaluated the impact of OER on academic outcomes [@Colvard2018], have been criticized for not including the information about the traditional and open textbooks used by the courses included in their studies [@Griggs2017]. We addressed this issue by retrieving the names of the textbooks from the relevant course syllabi. Prior to the adoption of the history open textbook [@SLCCHist1], [@foner2013give] was the textbook used introductory history courses. 

We also inspected the basic course structure before and after the open textbook adoption. Coursework varied across instructors, and typically included an assortment of readings, impact papers, quizzes, and examinations (for e.g. 2-6 in class examinations, take home exams) before and after the open textbook adoption. 

# Results

The primary goal of this project was to examine the impact of using an open textbook on academic outcomes of students. To this end, we examined the outcomes from an introductory history course at SLCC which piloted an open textbook in Spring 2016, and adopted it across all sections by Fall 2016. We examined academic outcomes in history after controlling for student characteristics such as pell eligibility and registration status, prior achievement measure such as cumulative undergraduate GPA, course type (classroom vs. online). In addition, we also compared academic outcomes in introductory history with introductory courses in other social sciences disciplines at SLCC (economics and political science) that used conventional textbooks during the same period. Analyzing academic outcomes in these other social sciences disciplines during the same period of time allowed us to account for any periodic changes across similar social science disciplines that may have contributed to changes in academic outcomes.

As a metric of academic outcome, previous studies have often combined drop, failure, and withdrawals [@Colvard2018; @Feldstein2012; @Hilton2016], pass rates [@Fischer2015], in place of or in addition to course grades. In fact, these drop/failure/withdrawal or pass metrics are often derived from thresholding course grades based on established standards, for e.g., any grade less than a 1.3 (< C-) is considered a failing grade. In order to draw more nuanced inferences, we used course grade as the primary outcome of interest. Although we also examined completion rates of courses, we found that all but `r sum(SLCC_11_18_proc$finalGrade == "I")` student completed their respective courses. Given that the courses were all required courses, students were likely to complete them in order to proceed with their curriculum, which explained why course completion was at ceiling levels. Therefore, we did not conduct any further analyses using course completion.

We first examined the descriptives for course grade, across factors of interest followed by a mixed effects hierarchical regression analysis [@cohen2013applied] in order to estimate the impact of open textbook adoption on course grade. This analysis is generally conducted by comparing at least two models:

- _base model_ that contains all relevant factors that might influence the outcome (course grade) except for the variable of interest (textbook type)          
- _comparison model(s)_ that includes the variable of interest in addition to all factors in the base model. If the comparison model significantly outperforms the base model, then we may safely conclude that the treatment had a consequential impact on the outcome. For the purposed of our analysis, we constructed two comparison models, an additive model, and an interactive model which will be explained in the following section.        

In our base and comparison models, the outcome variable was the grade each student received for each course s/he took. In addition to the outcome variable, we included the following _fixed effect_ predictor variables in the base model:

* student characterisitic variables such as whether the students were:
  + pell eligible          
  + registered as a part-time student         
* prior performance measures
  + cumulative undergraduate GPA           
  + Accuplacer reading comprehension score (20-120)
* course type, i.e., whether the course was delivered online  

Additionally, we included the following _random effect_ variables in the model:

* instructors nested under course subject accounting for instructor differences within each subject  
* given that course related changes often occur in annual or semesterly cycles, we accounted for these temporal factors by including the following temporal variables,
  + academic year (Fall to Summer constituting one academic year)         
  + semester (Fall, Summer, Spring)                    

Besides the base model, we constructed 2 comparison models: an additive, and an interactive model. The additive model included open-textbook type as the variable of interest. In the interactive model, in addition to open-textbook type, we included a term that accounted for the differential impact, if any, of open-textbook type on students who were pell eligible. The findings of the model results are represented in `r tbls("grade mod", display = "cite")`.


```{r grade mod, results='asis'}
stargazer::stargazer(gradeM0, gradeM1, gradeM2, type = "latex",omit.stat = c("BIC","ll"), covariate.labels = c("Pell Eligible","Registration Status (Part-Time)","Cumulative Undergraduate GPA","Accuplacer Reading Comprehension","Course Type (Online)","Textbook Type (Open) ", "Open Textbook x Pell Eligible"), dep.var.labels = "Course Grade Models", ci = TRUE, title = "Summaries for the base, additive, and interactive models, with course grade as the outcome measure.", dep.var.caption = "", digits = 2, column.labels = c("Base", "Additive", "Interactive"), model.numbers = FALSE, notes.align = "l", notes.label = "Note: 95\\% CI in parentheses", star.cutoffs = c(.05, .01, .001))
```

From the results of the base model `r tbls("grade mod",display = "cite")`, we observe that pell eligibility, cumulative undergraduate GPA, accuplacer reading comprehension scores, had a significant impact on course grade, while part-time registration status and online courses did not have a significant impact. Pell eligible students were likely to perform slightly worse than students who were not pell eligible. Furthermore, students with higher cumulative GPA, as well as students with  higher accuplacer reading comprehension score, had higher course grades. These factors retained the same impact across all three models (base, additive, interactive). From the additive and interactive models we found that open-textbook type had no significant effect on course grade. Likewise, in the interactive model we found that there was no significant difference in course grades for pell eligible students using open-textbooks. On comparing all the three models, we found that the base model had the least Akaike Information Criterion (AIC; ADD REF) value, and is therefore found to be the best out of all three models. Additionally, comparing the base model with the additive (_$\chi$_ (`r gradeComp$"Chi Df"[2]`) = `r gradeComp$Chisq[2]`, _p_ = `r gradeComp$"Pr(>Chisq)"[2]`) and interactive models (_$\chi$_ (`r gradeComp1$"Chi Df"[2]`) = `r gradeComp1$Chisq[2]`, _p_ = `r gradeComp1$"Pr(>Chisq)"[2]`) showed that these comparison models were not significantly different from the base model. Therefore, the comparison models did not outperform the base model.

While the model results clearly point to no significant effect of open-textbook type on course grade, this null effect is not captured when looking at the unadjusted means `r figs("nMasterPlot", display = "cite")`. In fact, inspection of the unadjusted course grade means might mislead readers into assuming a negative impact of OER on course grade. However, when we account for imbalances in the data by using the adjusted means (estimated marginal means), we observe that the null effect from our model results is corroborated by these adjusted means. We find further converging evidence of no significant effect of open-textbook use on course grade from the very small effect sizes (Cohen's d) obtained for variables of interest, contrasting groups using open-textbooks vs. those using traditional textbooks (`r tbls("effSizes", display = "cite")`; Cohen's d of greater than .20 and less than 0.5 is considered a small effect size). 

Therefore, from a linear mixed model hierarchical regression analysis we found that using open textbook did not have a significant impact on the course grades for the SLCC introductory history course. In the following section we discuss possible causes for this null effect, and implications for oer and future research.

```{r mMasterPlot, fig.cap="Unadjusted and adjusted history course grade means for non-pell eligible and pell eligible students."}
mMasterPlot
```


```{r effsizes}
effS %>% 
    kable(., format = "latex", linesep = "",caption = paste0("Effect size estimates (Cohen's d) showing non-significant group differences between students using traditional and open textbooks in the introductory history course at SLCC."), booktabs = TRUE,
          col.names = c("","Cohen's d[note]", "95% Conf. Int."),
          align = c("l", rep("r", 3))) %>%
    kable_styling(full_width = FALSE) %>%
    kableExtra::add_footnote(label = c("Standardized mean difference."),notation = "symbol") #95% confidence interval in brackets.
```

# References
<!--# Discussion

In a turbulent economy in which higher education has become a source of life-long financial burden, open education resources (OERs) are making a small but steady contribution to alleviating some of the financial burdens imposed on students. In addition to mitigating some of the monetary pressures, it is also esuring that students have access to study materials right when the semester starts, and no longer have to wait for textbooks to arrive weeks after the beginning of classes. While efforts in OER primarily focused on relieving financial stress, and ensuring and expanding access, the scientific community is now redirecting their focus on assessing the influence of OERs on educational outcomes, in an effort to examine how OERs compare with existing expensive closed resources. As this line of research is still in its infancy, the verdict on the impact of OER on academic outcomes is highly contested[@Hilton2016]. While some studies have shown a positive impact of OER [[ADD REF]], others have found a negative impact [[ADD REF]], and few have reported no effect of OER on academic outcomes.

In an effort to resolve this debate, we evaluated the impact of OER on course grades of an introductory history course at SLCC. We accounted for demographic factors, prior achievement measure, instructor level differences, and course type in order to isolate the impact of the open textbook on course grades. We found no significant effect of the open textbook adoption on grades, i.e., course grades for groups of students using an open textbook and a traditional textbook were similar. While this is a null effect, we can still appreciate that OER results in similar academic performance as when a traditional textbook is used. In the following section we speculate why we found no effect of OER on academic outcomes.

As previously mentioned, significant OER efforts are directed towards mitigating financial stress and improve access to educational materials. Therefore, for an OER adoption to result in positive academic outcomes, it needs to resolve situations in which a financial deterrent prevents students from buying a textbook for a course (e.g. textbook is very expensive), and/ or that students do not have any means to access the course textbook and would have to learn without it. In the case of the introductory history textbook however, we found that neither of these issues existed. An analysis of the course syllabi revealed that the price of previous textbook was \$25 (included in the tuition fees), and the textbook was available in a digital format on the institutional LMS from the beginning of the course. Students had the option of buying a printed copy of the textbook for another \$25. The cost of the open teextbook was \$5 (also included in the fees), and it was also available on the institutional LMS at the beginning of the semester. Given that the traditional textbook used was relatively inexpensive to begin with, and it was automatically available to students as a part of enrolling in the course, we can assume that as far as access and financial liability is concerned, the traditional textbook was characteristically similar to an open textbook.

 The findings of this study should not dishearten OER advocates, but allow them to understand that benefits of OER is most likely to be conspicuous in situations where the accessibility deficit exists. In situations where there is no discernible accessibility deficitgive them a sense of appreciation  While oer has no doubt made great strides in improving learning conditions for students in higher education

Therefore, adopting an open textbook likely did not increase access to the textbook, which contributed so no statistically detectable difference in course grades of students using traditional textbook relative to students using an open textbook. -->